---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
<!-- Variables marked with double curly braces will be transformed beforehand: -->
<!-- `packagename` is extracted from the DESCRIPTION file -->
<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cleanepi

**cleanepi** provides functions to clean epidemiological data provided in the form of a data frame or other related data type.   

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/cleanepi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/cleanepi/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/epiverse-trace/cleanepi/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/cleanepi?branch=main)
[![lifecycle-concept](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-concept.svg)](https://www.reconverse.org/lifecycle.html#concept) 
<!-- badges: end -->


## Installation

You can install the development version of cleanepi from
[GitHub](https://github.com/) with:

```{r}
# install.packages("remotes")
# remotes::install_github("epiverse-trace/cleanepi@develop", 
#                         build_vignettes=TRUE)
library(cleanepi)
```

## Manual  
```{r eval=FALSE}
browseVignettes("cleanepi")
```

## Description
| function name | description |    
| :-------------- | :---------- |   
| **clean_data** | cleaning of column names, detecting and removing empty rows and columns as well as constant columns and duplicated rows. It also replaces missing values by NA and automatically convert date columns |      
| **check_subject_ids** | check whether the IDs comply with the expected format |    
| **standardize_date** | convert date column into _%Y-%m-%d_ |     
| **calculate_age** | calculate age from date column. Returned age can be in either _years_, or _months_, or _weeks_, or _days_ or a combination of some of these|    
| **check_date_sequence** | check whether the sequence of event dates is correct |
| **WIP** | work in progress |


## Next steps
* build function to display the cleaning report
* build function to perform dictionary based cleaning


### Lifecycle

This package is currently a *concept*, as defined by the [RECON software
lifecycle](https://www.reconverse.org/lifecycle.html). This means that essential
features and mechanisms are still being developed, and the package is not ready
for use outside of the development team.


### Contributions

Contributions are welcome via [pull requests](https://github.com/{{ gh_repo }}/pulls).

### Code of Conduct

Please note that the cleanepi project is released with a 
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

