---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->

<!-- Variables marked with double curly braces will be transformed beforehand: -->

<!-- `packagename` is extracted from the DESCRIPTION file -->

<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cleanepi: Clean and standardize epidemiological data

<!-- badges: start -->

[![License:
MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/cleanepi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/cleanepi/actions/workflows/R-CMD-check.yaml)
[![Codecov test
coverage](https://codecov.io/gh/epiverse-trace/cleanepi/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/cleanepi?branch=main)
[![lifecycle-concept](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-concept.svg)](https://www.reconverse.org/lifecycle.html#concept)

<!-- badges: end -->

**cleanepi** is an R package to clean, curate, and standardize epidemiological data. 

**cleanepi**  performs several data cleaning tasks that an end-user would anticipate to be performed on a cluttered dataset. These include standard tasks such as removing duplicated and empty rows and row, detecting columns of constant value, replacing missing values, performing a sanity check on uniquely identified columns, and converting dates in a string to date-format. **cleanepi** can also standardize column entries into specified formats, calculate age (in years, months, weeks, or days) for a given column and reference data, and perform further dictionary-based cleaning operations.
**cleanepi** expects input data  in  form of  dataframe-like structure (dataframe, tibble, and table) or linelist object, and returns dataframe containing cleaned data  and a report detailing the carried-out operations. 

**cleanepi** is developed by  the [Epiverse-TRACE](https://data.org/initiatives/epiverse/) team at the Medical Research Council The Gambia unit at the London School of Hygiene and Tropical Medicine (MRCG@LSHTM).


## Installation

The current development version of **cleanepi** can be installed from  [here](https://epiverse-trace.github.io/cleanepi/dev/).

```{r}
# install.packages("remotes")
# remotes::install_github("epiverse-trace/cleanepi@develop", 
#                         build_vignettes=TRUE)
library(cleanepi)
```

## Quick start
The main function in **cleanepi** is `clean_data(),` which calls 
functions for almost all standard data cleaning tasks, such as removal of empty and duplicated rows and columns, replacement of missing values, etc. However, each function can also be called independently to perform a specific  task.

## Description
| function name | description |    
| :-------------- | :---------- |   
| **clean_data** | cleaning of column names, detecting and removing empty rows and columns as well as constant columns and duplicated rows. It also replaces missing values by NA and automatically convert date columns |      
| **check_subject_ids** | check whether the IDs comply with the expected format |    
| **standardize_date** | convert date column into _%Y-%m-%d_ |     
| **calculate_age** | calculate age from date column. Returned age can be in either _years_, or _months_, or _weeks_, or _days_ or a combination of some of these|    
| **check_date_sequence** | check whether the sequence of event dates is correct |
| **find_duplicates** | detect and return duplicated rows from the input dataset |
| **remove_duplicates** | remove detected duplicated rows |
| **WIP** | work in progress |

`clean_data()` returns a `list` containing `data.frame` for the  processed input data and a `list` of reports for each operation. Below is typical example of how to use `clean_data()` function. 

In this example, the following data cleaning operations are performed on the input dataset based on the following criteria:  

1. removal of any duplicated line across all columns (`remove_duplicates = TRUE`; `target_columns = NULL`)
2. replacement of the missing character `-99` by `NA` (`replace_missing = TRUE`; `na_comes_as = "-99"`)
3. conversion of any `character` column into `Date` if the 50% or more of its values turn out to be dates. Dates not within the specified time frame will be considered as outliers and replaced by `NA` (`check_timeframe = TRUE`; `timeframe = as.Date(c("1973-05-29", "2023-05-29"))`; `error_tolerance = 0.5`)
4. Checking if the subject IDs comply with the expected format (`subject_id_col_name = "study_id"`; `subject_id_format = "PS000P2"`; `prefix = "PS"`; `suffix = "P2"`; `range = c(1, 100)`). This will detect and remove the rows where the format of the subject IDs are incorrect.
5. The other cleaning operations, mentioned above, will be automatically applied to the input dataset. 


```{r eval=TRUE}
# READING IN THE TEST DATASET
test_data <- readRDS(system.file("extdata", "test_df.RDS", 
                                 package = "cleanepi"))

# VISUALISE THE INPUT DATASET
print(test_data)

# DEFINING THE CLEANING PARAMETERS
params <- list(
  remove_duplicates = TRUE,
  target_columns = NULL,
  replace_missing = TRUE,
  na_comes_as = "-99",
  check_timeframe = TRUE,
  timeframe = as.Date(c("1973-05-29", "2023-05-29")),
  error_tolerance = 0.5,
  subject_id_col_name = "study_id",
  subject_id_format = "PS000P2",
  prefix = "PS",
  suffix = "P2",
  range = c(1, 100)
  )

# PERFORMING THE DATA CLEANING
res <- clean_data(
  data = test_data,
  params = params
)

cleaned_data <- res$data
cleaning_report <- res$report

# VISUALISE THE CLEANED DATASET
print(cleaned_data)
```

## Next steps

* build function to display the cleaning report
* build function to perform dictionary based cleaning
* build function to quantify and handle missing data


### Lifecycle

This package is currently a *concept*, as defined by the [RECON software
lifecycle](https://www.reconverse.org/lifecycle.html). This means that
essential features and mechanisms are still being developed, and the
package is not ready for use outside of the development team.

### Contributions

Contributions are welcome via [pull
requests](https://github.com/%7B%7B%20gh_repo%20%7D%7D/pulls).

### Code of Conduct

Please note that the cleanepi project is released with a [Contributor
Code of
Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
