---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
<!-- Variables marked with double curly braces will be transformed beforehand: -->
<!-- `packagename` is extracted from the DESCRIPTION file -->
<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cleanepi

**cleanepi** provides functions to clean epidemiological data provided in the form of a data frame or other related data type.   

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/cleanepi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/cleanepi/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/epiverse-trace/cleanepi/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/cleanepi?branch=main)
[![lifecycle-concept](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-concept.svg)](https://www.reconverse.org/lifecycle.html#concept) 
<!-- badges: end -->


## Installation

You can install the development version of cleanepi from
[GitHub](https://github.com/) with:

```{r}
# install.packages("remotes")
# remotes::install_github("epiverse-trace/cleanepi", build_vignettes=TRUE)
library(cleanepi)
```

## Manual  
```{r}
browseVignettes("cleanepi")
```

## Description
| function name | description |    
| :-------------- | :---------- |   
| **cleanepi** | check whether the IDs comply with the expected format. It will check for: mistakes in subject IDs prefix and suffix, subject IDs with incorrect length, number in subject IDs that are out of range or not permitted |      
| [check_subject_ids](./doc/check_subject_ids.md) | check whether the IDs comply with the expected format |    
| **standardize_date** | convert date column into _%Y-%m-%d_ |     
| **calculate_age** | calculate age from date column. Returned age can be in either _years_, or _months_, or _weeks_, or _days_ or a combination of some of these|    
| **check date sequence** | check the sequence of event dates |
| **WIP** | work in progress |


## Development

### Lifecycle

This package is currently a *concept*, as defined by the [RECON software
lifecycle](https://www.reconverse.org/lifecycle.html). This means that essential
features and mechanisms are still being developed, and the package is not ready
for use outside of the development team.


### Contributions

Contributions are welcome via [pull requests](https://github.com/{{ gh_repo }}/pulls).

Contributors to the project include:

* Karim Mané 
* 

### Code of Conduct

Please note that the cleanepi project is released with a 
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.

